---
title: "ProFit: Multi-frame and Multi-band Fitting"
author: "Aaron Robotham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProFit: Multi-frame and Multi-band Fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load our useful libraries:

```{r}
library(ProFit)
library(ProFound)
library(magicaxis)
library(Highlander)
library(Rfits)
```

Set global evaluate (basically TRUE for GitHub version and FALSE for CRAN):

```{r}
evalglobal = TRUE
```

## Fitting Simple Offset Images

Here we look at co-fitting images with offset positions.

```{r}
im1 = Rfits_read_image(system.file("extdata/Offset/Zim1.fits",package="ProFit"), ext=2)$imDat
im2 = Rfits_read_image(system.file("extdata/Offset/Zim2.fits",package="ProFit"), ext=2)$imDat
```

Here we have two similar Z-band frames taken at slightly offset positions of the same primary source. With **ProFit** it is possible to fit these simultaneously without resorting to stacking.

```{r fig.width=5, fig.height=5, dpi=100}
magimage(im1)
magimage(im2)
```

First we should extract the PSFs using the **profitAllStarDoFit** function (this will take about 5 minutes):

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
PSF1 = profitAllStarDoFit(im1, magzero=30, rough=TRUE, plot=TRUE)
```

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
PSF2 = profitAllStarDoFit(im2, magzero=30, rough=TRUE, plot=TRUE)
```

We can compare the models:

```{r eval=evalglobal}
PSF1$psf_modellist
PSF2$psf_modellist
```

The seeing in our second image is clearly quite a bit better (FWHM 2.4 versus 2.9). This is also why we should get a better solution by fitting both images simultaneously rather than stacking.

To get most of the way to the correct Datalist structure we need we can use **profitFound2Fit**:

```{r eval=evalglobal}
F2F1 = profitFound2Fit(im1, loc=c(450,500), psf=PSF1$psf, cutbox=c(300,300), Ncomp=2,
                       magzero=30, SBdilate=2, tightcrop=FALSE)
F2F2 = profitFound2Fit(im2, loc=c(450,500), psf=PSF2$psf, cutbox=c(300,300), Ncomp=2,
                       magzero=30, SBdilate=2, tightcrop=FALSE)
```

The main adjustment we need to make it to tell **ProFit** what the offset is between the frames and put this into the *Data* object contained. From the original WCSs the offset from *im1* to *im2* is know to be [-23.112,-9.697] pixels:

```{r eval=evalglobal}
F2F2$Data$offset = c(-23.112,-9.697)
```

We can now check this using the **profitLikeModel** function, where the same initial parameters get shifted for the second *Data* object:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
profitLikeModel(F2F1$Data$init, Data=F2F1$Data, makeplots=TRUE)
profitLikeModel(F2F1$Data$init, Data=F2F2$Data, makeplots=TRUE)
```
The solutions look pretty similar now! To simplify things we want to create a *Datalist* structure that contains the two lists:

```{r eval=evalglobal}
Datalist = c(list(F2F1$Data), list(F2F2$Data))
```

You can now run **profitLikeModel** more directly:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
profitLikeModel(F2F1$Data$init, Data=Datalist, makeplots=TRUE)
```
The final thing we need to do is add a couple of elements to the *Datalist* to make it compatible with **LaplacesDemon**:

```{r eval=evalglobal}
Datalist$mon.names = F2F1$Data$mon.names
Datalist$parm.names = F2F1$Data$parm.names
Datalist$N = F2F1$Data$N
```

Now we can finally run **Highlander** to find the best common solution for the two frames (this will take about 5 minutes):

```{r eval=evalglobal}
HighFit = Highlander(F2F1$Data$init, Data=Datalist, likefunc=profitLikeModel, ablim=1)
```

And now we can check the results:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
profitLikeModel(HighFit$parm, Data=Datalist, makeplots=TRUE)
```

A similar methodology will work on any number of images, as long as they fit within memory. They should strictly all be the same band and pixel scale however, since currently (as of Jan 2021) **ProFit** does not allow the main model parameters to vary between images (other than the [X,Y] offset as shown above). Longer term the plan is to allow such functionality.

## Free, Fixed, and Follow Fits

It is also possible, but more complicated, to fit some but not all parameters. For this mode the users will need to be keenly aware of which parameters are which within the parm stuctures of *Data* since they will need to construct their own, and tag which parts should correspond to the relevant model list. If that sounds complicated, then its because it is! With great flexibility comes great complexity...

Imagine we want to do a fit much like the above, but where we want to the two disk Re argument to be free to fit differently in image 2. In this case the image is of the same object in the same band, so there is no real reason to allow the disk Re to fit freely (they should be linked together) but we can treat it as a placeholder for an observation in a different band. In this case we might expect many parameters to be precisely linked ([X,Y] and position angle), but the size (Re) to be different because we expect colour gradients in (at least some) disks. As a simple example we can look at the above *Datalist*.

```{r eval=evalglobal}
Datalist[[1]]$init
Datalist[[2]]$init
```

We now need to make a new *init* structure that adds on an additional re term for the second image disk (re2b) and a new *parmuse* structure that tells **ProFit** where to look for parameters when fittin each image.

```{r eval=evalglobal}
Datalist2 = c(list(F2F1$Data), list(F2F2$Data)) #Note we are still using the offset added to F2F2$Data

init_new = F2F1$Data$init
init_new = c(init_new, sersic.re2b = as.numeric(F2F2$Data$init['sersic.re2']))
init_new

Datalist2$mon.names = F2F1$Data$mon.names
Datalist2$parm.names = names(init_new) #Careful: This needs to reflect the new longer parm init.
Datalist2$N = F2F1$Data$N

Datalist2[[1]]$init = init_new
Datalist2[[2]]$init = init_new

Datalist2[[1]]$parmuse = 1:8 #We use the first arguments for image 1 fitting
Datalist2[[2]]$parmuse = c(1,2,3,4,5,9,7,8)
```

We can sanity check this by passing through **profitLikeModel** as before:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
profitLikeModel(Datalist2[[1]]$init, Data=Datalist2, makeplots=TRUE)
```

Now we can finally run **Highlander** to find the best common solution for the two frames (this will take about 5 minutes) where the disk Re are allowed to differ:

```{r eval=evalglobal}
HighFit2 = Highlander(Datalist2[[1]]$init, Data=Datalist2, likefunc=profitLikeModel, ablim=1)
```

We still a get a pretty good fit:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
profitLikeModel(HighFit2$parm, Data=Datalist2, makeplots=TRUE)
```

Now we can take a look at the fit parameters. Note that even though they were free to be fitted separately, sersic.re2 and sersic.re2b are very similar!

```{r eval=evalglobal}
HighFit2$parm
```

## Fully Automatic Multi-band Fitting

**ProFit** v2.0.0 and upwards support highly automated multi-band fitting via the **profitMultiBandFound2Fit** and **profitMultiBandDoFit** functions. These require the images provided to be exactly registered and cropped to the desired level (no offset or cut out options here). If this is not the case, then lower level functions (i.e. those above) probably need to be used instead. These functions will find stars and create per band PSFs automatically, so they might well covere the major use cases for many people.

As an example we can load in 9 bands of data that come included in the **ProFound** package, and cut out around a bright spiral galaxy:

```{r eval=evalglobal}
loc = c(1200,480)
cut = -299:300

image_list = list(
  u = Rfits_read_image(system.file("extdata", 'MultiBand/u.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  g = Rfits_read_image(system.file("extdata", 'MultiBand/g.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  r = Rfits_read_image(system.file("extdata", 'MultiBand/r.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  i = Rfits_read_image(system.file("extdata", 'MultiBand/i.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  Z = Rfits_read_image(system.file("extdata", 'MultiBand/Z.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  Y = Rfits_read_image(system.file("extdata", 'MultiBand/Y.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  J = Rfits_read_image(system.file("extdata", 'MultiBand/J.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  H = Rfits_read_image(system.file("extdata", 'MultiBand/H.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut],
  Ks = Rfits_read_image(system.file("extdata", 'MultiBand/Ks.fits', package="ProFound"),ext=2)$imDat[loc[1] + cut, loc[2] + cut]
)
```

We wish to force some degree of smoothness in certain per-band parameters (bulge mag [mag1] disk mag [mag2] and disk Re [re2]), so we make use of the *smooth.parm* function argument where we limit the smoothness to having 3 degrees of freedom:

```{r eval=evalglobal}
smooth.parm = list(
  mag1=function(x,y){smooth.spline(x,y,df=3)},
  mag2=function(x,y){smooth.spline(x,y,df=3)},
  re2=function(x,y){smooth.spline(x,y,df=3)}
)
```

By supplying these functions we optimise the node points in our spline rather than the parameters directly. This hopefully means we reduce unphysical discontinuities, whilst still achieving a good quality fit.

We also need to tell it the wavelength of the images provided, listed in blue to red order:

```{r eval=evalglobal}
wave = c(pivwave[grep('VST', pivwave$filter)[1:4],"pivwave"], #VST u/g/r/i
         pivwave[grep('VISTA', pivwave$filter)[1:5],"pivwave"]) #VISTA Z/Y/J/H/Ks
```

We can now run a highly automated fit, where the only significant non-default option is to run the stacked image **profoundProFound** run using a tolerance of 20 to stop the target large spiral galaxy being broken up. This will take about 45 minutes to run (go have a nice coffee break):

```{r eval=evalglobal}
MultiFit = profitMultiBandDoFit(image_list=image_list, magzero=c(0,0,0,0,30,30,30,30,30),
                                smooth.parm=smooth.parm, wave=wave, tolerance=20)
```

We can check the transition from disk to bulge dominated galaxy light as we move from bluer to redder bands:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
bulge_mag = MultiFit$parm_smooth[grep('mag1',names(MultiFit$parm_smooth))]
disk_mag = MultiFit$parm_smooth[grep('mag2',names(MultiFit$parm_smooth))]

magplot(wave, bulge_mag, ylim=c(19,15.5), type='l', col='red', log='x', xlab='Wave / Ang', ylab='Mag')
lines(wave, disk_mag, col='blue')
legend('bottomright', legend=c('bulge', 'disk'), lty=1, col=c('red','blue'))
```

It is often instructive to show the above as the evolution of B/T (bulge flux divided by total flux):

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
magplot(wave,10^(-0.4*bulge_mag) / (10^(-0.4*bulge_mag) + 10^(-0.4*disk_mag)), ylim=c(0,1), type='l', log='x', xlab='Wave / Ang', ylab='B/T')
abline(h=0.5, lty=2)
```

We can also compare the disk Re to the bulge. We can see that the disk Re drops as a function of wavelength:

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
magplot(wave,10^MultiFit$parm_smooth[grep('re2',names(MultiFit$parm_smooth))], ylim=c(0,80), type='l', col='blue', log='x', xlab='Wave / Ang', ylab='Re / px')
abline(h=10^MultiFit$parm_smooth["sersic.re1"], col='red')
legend('bottomright', legend=c('bulge', 'disk'), lty=1, col=c('red','blue'))
```

We can also view all of the individual per-band fits (order blue-red ugriZYJHKs):

```{r fig.width=5, fig.height=3, dpi=100, eval=evalglobal}
profitLikeModel(MultiFit$parm, MultiFit$Data_list, makeplots=TRUE) #use MultiFit$parm, since profitLikeModel applies the smoothing
```
